---
title: "differential expression analysis"
author: "Manuel Cortes"
date: "2023-09-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# if (!require("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")

# BiocManager::install("clusterProfiler")
# BiocManager::install("biomaRt")

# Attach the DESeq2 library
library(DESeq2)

# Attach the ggplot2 library for plotting
library(ggplot2)

# We will need this so we can use the pipe: %>%
library(magrittr)

# Human Annotation
library(org.Hs.eg.db)


library(biomaRt)  
library(clusterProfiler)


set.seed(123)


```

# Analysis

# Please see differential-expression-analysis.Rmd to load all necessary variables
# Here, we will simply load the result from it

#First we will look at the top 5000 genes with the most variance.


```{r}
# Calculate variance for each gene (row-wise)
gene_variance <- apply(expression_df, 1, var)

# Sort genes by variance in descending order and get the top 5000
top_genes_indices <- order(gene_variance, decreasing = TRUE)[1:5000]

# Extract the top 5000 genes
top_genes_df <- expression_df[top_genes_indices,]

# Standardize the Data
scaled_matrix <- (scale(t(top_genes_df))) # Transpose is needed because scale operates on columns
```

```{r}

set.seed(123)

# K-means clustering
kmeans_result <- kmeans(scaled_matrix, centers=4)

# Visualization using PCA
pca_result <- prcomp(scaled_matrix)
plot(pca_result$x[,1:2], col=kmeans_result$cluster, pch=16, 
     xlab="PC1", ylab="PC2", main="K-means Clustering of Top 5000 Genes")

```
# This is for analyzing the number of clusters to use for k means. the analysis will differ for different clustering algorithms

```{r}
# Compute and plot wss for k = 1 to k = 10
wss <- sapply(1:10, function(k){
  kmeans(scaled_matrix, centers=k)$tot.withinss
})

plot(1:10, wss, type="b", pch=19, frame=FALSE, 
     xlab="Number of clusters K", 
     ylab="Total within-clusters sum of squares")

```

```{r}
# 1. Merge clustering results with metadata
# Assuming the row names of scaled_matrix are the SAMPLE_NAME
result_df <- data.frame(refinebio_accession_code = rownames(scaled_matrix), 
                        Cluster = kmeans_result$cluster)
merged_df <- merge(result_df, metadata, by="refinebio_accession_code")

# Order merged_df by the order of samples in PCA result
merged_df <- merged_df[match(rownames(pca_result$x), merged_df$refinebio_accession_code), ]

# 2. Visualize the PCA results colored by clusters but shaped by metadata GROUP
# # Convert the GROUP column to a factor for using different plotting symbols
merged_df$Group <- as.factor(merged_df$Group)

ggplot(data = merged_df, aes(x = pca_result$x[,1], y = pca_result$x[,2], color = as.factor(Cluster), shape = Group)) +
  geom_point(size = 3) +
  scale_shape_manual(values = c(0, 1, 2, 3, 4, 5, 6, 7, 8, 10, 15, 17)) +
  labs(title = "K-means Clustering vs Predefined Groups", x = "PC1", y = "PC2") +
  theme_minimal()


```

```{r}
# Define the numbers of top genes we want to use for clustering
top_genes_numbers <- c(10, 100, 1000, 5000, 10000)

# Calculate variance for each gene (row-wise)
gene_variance <- apply(expression_df, 1, var)

# Sort genes by variance in descending order
sorted_genes_indices <- order(gene_variance, decreasing = TRUE)

# Initialize an empty list to store the kmeans results
kmeans_results <- list()

set.seed(123)

# Iterate over each top_genes_number and perform clustering
for (n in top_genes_numbers) {
  
  # Extract the top genes
  top_genes_df <- expression_df[sorted_genes_indices[1:n], ]
  
  # Standardize the Data
  scaled_matrix <- (scale(t(top_genes_df))) # Transpose is needed because scale operates on columns
  
  # K-means clustering
  kmeans_results[[paste0("Top", n)]] <- kmeans(scaled_matrix, centers=4)
}

# Now, kmeans_results is a list with each element being the result of kmeans for a particular number of top genes

```

```{r}
# Initialize an empty data frame to store the clustering results
alluvial_data <- data.frame(Sample = colnames(expression_df))

# Iterate over each clustering result in kmeans_results and add it to the data frame
for (name in names(kmeans_results)) {
  alluvial_data[[name]] <- kmeans_results[[name]]$cluster
}

# View the first few rows to ensure it looks correct
head(alluvial_data)

```
```{r}
library(ggplot2)
library(ggalluvial)

ggplot(data = alluvial_data, 
       aes(axis1 = Top10, axis2 = Top100, axis3 = Top1000, axis4 = Top5000, axis5 = Top10000)) +
  geom_alluvium(aes(fill = factor(Top10))) +
  geom_stratum() +
  geom_text(stat = "stratum", aes(label = after_stat(stratum))) +
  theme_minimal() +
  labs(title = "Changes in Cluster Memberships Across Different Clustering Setups")

```

```{r}

# Create contingency table
contingency_table <- table(merged_df$Cluster, merged_df$Group)

# Removing the 'Other' column
contingency_table <- contingency_table[, -which(colnames(contingency_table) == "Other")]


# Perform chi-squared test
test_result <- chisq.test(contingency_table)

# Store p-value
p_value <- test_result$p.value

ggplot(data = as.data.frame(as.table(contingency_table)), aes(Var1, Var2)) +
  geom_tile(aes(fill = Freq), color = "white") +
  geom_text(aes(label = sprintf("%d", Freq)), vjust = 1) +
  scale_fill_gradient(low = "white", high = "steelblue") +
  theme_minimal() +
  labs(x = "Cluster", y = "Group", fill = "Count")


contingency_table

```

```{r}
# Define the number of genes for each iteration
num_genes <- c(10, 100, 1000, 5000, 10000)

# Initialize an empty list to store the kmeans results
kmeans_results <- list()

for (genes in num_genes) {
  # Extract top genes by variance
  top_genes_indices <- order(gene_variance, decreasing = TRUE)[1:genes]
  top_genes_df <- expression_df[top_genes_indices,]

  # Standardize the Data
  scaled_matrix <- (scale(t(top_genes_df)))

  # K-means clustering
  set.seed(123)
  kmeans_results[[as.character(genes)]] <- kmeans(scaled_matrix, centers=4)
}

# Initialize vectors to store p-values
p_values <- numeric()
adjusted_p_values <- numeric()

for (genes in names(kmeans_results)) {
  # Merge with metadata
  result_df <- data.frame(refinebio_accession_code = rownames(scaled_matrix),
                          Cluster = kmeans_results[[genes]]$cluster)
  merged_df <- merge(result_df, metadata, by="refinebio_accession_code")

  # Create contingency table and perform chi-squared test
  contingency_table <- table(merged_df$Cluster, merged_df$Group)
  contingency_table <- contingency_table[, -which(colnames(contingency_table) == "Other")]

  test_result <- chisq.test(contingency_table)

  # Store p-values
  p_values <- c(p_values, test_result$p.value)
}

# Adjust p-values for multiple hypothesis testing
adjusted_p_values <- p.adjust(p_values, method = "BH")

results_table <- data.frame(Number_of_Genes = num_genes,
                            Unadjusted_P_Value = p_values,
                            Adjusted_P_Value = adjusted_p_values)

print(results_table)
```
#Heatmap generation

```{r}

# Assuming you have the sample groups from Assignment 1 in a vector called sample_groups
annotations <- data.frame(
  KMeans = merged_df$Cluster,
  # You should add other clustering methods here.
  SampleGroup = merged_df$Group # Placeholder. Replace with actual data.
)

annotations$KMeans <- as.factor(annotations$KMeans)

library(ComplexHeatmap)

Heatmap(
  matrix = t(scaled_matrix), # Transposed so genes are rows and samples are columns
  name = "expression",       # Name of the heatmap shown in the legend
  show_row_names = FALSE,    # Might not want to display 5000 gene names
  show_column_names = FALSE, # Depends on your sample size
  cluster_rows = TRUE,       # Cluster genes
  cluster_columns = TRUE,    # Cluster samples
  top_annotation = HeatmapAnnotation(df = annotations)
)


```


```{r}



```
# Misc DE Stuff

# This part really is not needed for this assignment
```{r}

# Define the file path
file_path <- file.path("results", "SRP076307_diff_expr_results_ensembl.tsv")

# Read the file
deseq_df <- readr::read_tsv(file_path)


# # Filter for significantly differentially expressed genes using the threshold column
# sig_genes_df <- deseq_df %>%
#   dplyr::filter(threshold == TRUE)
# entrez_ids <- sig_genes_df$Gene

# Sorting by the absolute value of log2FoldChange in descending order
top_de_genes <- deseq_df %>%
  dplyr::arrange(dplyr::desc(abs(log2FoldChange))) %>%
  dplyr::slice_head(n = 5000)

# Extracting gene IDs from the top differentially expressed genes
top_de_genes_id <- top_de_genes$Gene

# Subsetting the main expression matrix to only include rows for the top differentially expressed genes using row names
top_de_genes_df <- expression_df[rownames(expression_df) %in% top_de_genes_id,]
# top_de_genes_df <- gene_matrix[top_de_genes_id,]

# Standardizing (scaling) the expression data for top differentially expressed genes
scaled_matrix_de <- t(scale(top_de_genes_df))

```

# Clustering analysis with K means

```{r}

kmeans_result_de <- kmeans(scaled_matrix_de, centers = 4)

pca_result_de <- prcomp(scaled_matrix_de, scale.=FALSE)
plot(pca_result_de$x[,1:2], col=kmeans_result_de$cluster, pch = 16, xlab = "PC1", ylab = "PC2")

```
```{r}
# 1. Merge clustering results with metadata
# Assuming the row names of scaled_matrix are the SAMPLE_NAME
result_df <- data.frame(refinebio_accession_code = rownames(scaled_matrix_de), 
                        Cluster = kmeans_result_de$cluster)
merged_df <- merge(result_df, metadata, by="refinebio_accession_code")

# Order merged_df by the order of samples in PCA result
merged_df <- merged_df[match(rownames(pca_result$x), merged_df$refinebio_accession_code), ]

# 2. Visualize the PCA results colored by clusters but shaped by metadata GROUP
# # Convert the GROUP column to a factor for using different plotting symbols
merged_df$Group <- as.factor(merged_df$Group)

ggplot(data = merged_df, aes(x = pca_result$x[,1], y = pca_result$x[,2], color = as.factor(Cluster), shape = Group)) +
  geom_point(size = 3) +
  scale_shape_manual(values = c(0, 1, 2, 3, 4, 5, 6, 7, 8, 10, 15, 17)) +
  labs(title = "K-means Clustering vs Predefined Groups", x = "PC1", y = "PC2") +
  theme_minimal()

```
```{r}
# Compute and plot wss for k = 1 to k = 10
wss_de <- sapply(1:10, function(k){
  kmeans(scaled_matrix_de, centers=k)$tot.withinss
})

plot(1:10, wss_de, type="b", pch=19, frame=FALSE, 
     xlab="Number of clusters K", 
     ylab="Total within-clusters sum of squares")

```

# Trying out hclust

```{r}

# 1. Calculate the distance matrix
dist_matrix <- dist(scaled_matrix_de, method = "euclidean")

# 2. Perform Hierarchical Clustering
hclust_result <- hclust(dist_matrix, method = "complete")

# 3. Visualize the Dendrogram
plot(hclust_result)

# 4. Cut the dendrogram to form a specific number of clusters (e.g., 3 clusters)
clusters <- cutree(hclust_result, k = 4)

# 5. Visualize the clusters on the PCA plot (assuming `pca_result` was computed earlier)
plot(pca_result_de$x[,1:2], col=clusters, pch=16, xlab="PC1", ylab="PC2")

```
```{r}
library(pheatmap)

# Create the annotations data frame
annotations <- data.frame(
  Kmeans_Cluster = kmeans_result_de$cluster
  # Hierarchical_Cluster = hclust_result_clusters, # replace with actual data
  # Sample_Groups = sample_groups_assignment1      # replace with actual data
  # ... add other clustering results as needed
)

# Draw the heatmap
pheatmap(
  scaled_matrix_de,
  annotation_col = annotations,
  show_rownames = FALSE,
  show_colnames = TRUE,
  clustering_distance_rows = "euclidean",
  clustering_distance_cols = "euclidean",
  clustering_method = "complete",
  main = "Heatmap of Top 5,000 Genes"
)


```


```{r}
any(is.na(scaled_matrix_de))
any(is.infinite(scaled_matrix_de))
nrow(annotations) == ncol(scaled_matrix_de)


```
