---
title: "differential expression analysis"
author: "Manuel Cortes"
date: "2023-09-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# if (!require("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")

# BiocManager::install("clusterProfiler")
# BiocManager::install("biomaRt")

# Attach the DESeq2 library
library(DESeq2)

# Attach the ggplot2 library for plotting
library(ggplot2)

# We will need this so we can use the pipe: %>%
library(magrittr)

# Human Annotation
library(org.Hs.eg.db)


library(biomaRt)  
library(clusterProfiler)
# Load necessary libraries
library(tidymodels)
library(kernlab) # for the SVM

set.seed(123)


```

# Analysis

# Please see differential-expression-analysis.Rmd to load all necessary variables
# Here, we will simply load the result from it

#First we will look at the top 5000 genes with the most variance.

```{r}
# Calculate variance for each gene (row-wise)
gene_variance <- apply(expression_df, 1, var)

# Sort genes by variance in descending order and get the top 5000
top_genes_indices <- order(gene_variance, decreasing = TRUE)[1:5000]

# Extract the top 5000 genes
top_genes_df <- expression_df[top_genes_indices,]

# Standardize the Data
scaled_matrix <- (scale(t(top_genes_df))) # Transpose is needed because scale operates on columns
```

# This will run the classifier across all groups in our data for top 5000 genes

```{r}

# Assuming you have metadata with labels in a data frame called 'metadata_df'
# Ensure that the ordering matches the expression data
labels <- metadata$Group

# Prepare data for training
data_train <- as.data.frame(scaled_matrix) # No need to transpose
data_train$label <- as.factor(labels)

# Splitting data into training and testing sets (e.g., 80% training, 20% testing)
set.seed(123) # Set seed for reproducibility
data_split <- initial_split(data_train, prop = 0.8)
train_data <- training(data_split)
test_data <- testing(data_split)

# Define SVM model specification
svm_spec <- svm_poly() %>%
  set_engine("kernlab") %>%
  set_mode("classification")

# Train the model
svm_fit <- svm_spec %>%
  fit(label ~ ., data = train_data) # Train SVM using training data

# Make predictions and evaluate the model
predictions <- predict(svm_fit, test_data)

# Add predictions to the test data
test_data$predicted_label <- predictions$.pred_class

# Compute metrics
metrics_result <- yardstick::metrics(test_data, truth = label, estimate = predicted_label)
print(metrics_result)

```
# Same as above, but generalized to different numbers of genes

```{r}
# Preprocess data function
preprocess_data <- function(expression_df, n_genes) {
  gene_variance <- apply(expression_df, 1, var)
  top_genes_indices <- order(gene_variance, decreasing = TRUE)[1:n_genes]
  top_genes_df <- expression_df[top_genes_indices, ]
  scaled_matrix <- scale(t(top_genes_df))
  return(scaled_matrix)
}

# Analysis function
perform_analysis <- function(scaled_matrix, labels) {
  data_train <- as.data.frame(scaled_matrix)
  data_train$label <- as.factor(labels)
  
  set.seed(123)
  data_split <- initial_split(data_train, prop = 0.8)
  train_data <- training(data_split)
  test_data <- testing(data_split)
  
  svm_spec <- svm_poly() %>%
    set_engine("kernlab") %>%
    set_mode("classification")
  
  svm_fit <- svm_spec %>%
    fit(label ~ ., data = train_data)
  
  predictions <- predict(svm_fit, test_data)
  test_data$predicted_label <- predictions$.pred_class
  
  metrics_result <- yardstick::metrics(test_data, truth = label, estimate = predicted_label)
  
  print(paste("Results for top", ncol(scaled_matrix), "genes:"))
  print(metrics_result)
  
  # Return metrics and SVM weights
  list(metrics = metrics_result, svm_model = svm_fit$fit)
}

# Main loop
gene_counts <- c(10, 100, 1000, 5000)#, 10000)
labels <- metadata$Group

# Initialize vectors and lists
accuracies <- c()
svm_models_list <- list()

for (n_genes in gene_counts) {
  scaled_matrix <- preprocess_data(expression_df, n_genes)
  analysis_result <- perform_analysis(scaled_matrix, labels)
  
  # Extract accuracy and store in vector
  acc <- analysis_result$metrics$`.estimate`[analysis_result$metrics$`.metric` == "accuracy"]
  accuracies <- c(accuracies, acc)
  
  # Store SVM weights
  svm_models_list[[as.character(n_genes)]] <- analysis_result$svm_model
}

# Plotting Accuracy vs. Number of Genes
df <- data.frame(Number_of_Genes = gene_counts, Accuracy = accuracies)
ggplot(df, aes(x = Number_of_Genes, y = Accuracy)) +
  geom_line() +
  geom_point() +
  labs(title = "Accuracy vs. Number of Genes", x = "Number of Genes", y = "Accuracy") +
  theme_minimal()


```

# looking at Gene Signatures - THIS PART ISNT WORKING VERY WELL

```{r}

# Accessing the SVM model from the list
svm_model_fit <- svm_models_list[['1000']]

# Extract the support vectors
support_vectors <- svm_model_fit@xmatrix

# Number of support vectors
cat("Number of support vectors:", nrow(support_vectors), "\n")

# Extract the coefficients (weights)
weights <- svm_model_fit@coef

# Extract all weights and combine them
all_weights <- unlist(weights)

# Visualizing the distribution of the weights
df_weights <- data.frame(Weight = all_weights)
ggplot(df_weights, aes(x=Weight)) +
  geom_histogram(bins=30, fill="blue", alpha=0.7) +
  labs(title="Distribution of SVM Weights") +
  theme_minimal()

# # Visualizing weights for each SVM model (histogram for each number of genes)
# for (n_genes in gene_counts) {
#   # Extract weights
#   weights <- abs(unlist(svm_weights_list[[as.character(n_genes)]]))
#   
#   # Plot histogram
#   p <- ggplot(data.frame(Weights = weights), aes(x = Weights)) +
#     geom_histogram(binwidth = 0.01, fill = "blue", color = "black", alpha = 0.7) + 
#     labs(title = paste("Distribution of SVM Model Weights for top", n_genes, "genes"), x = "Weight Value", y = "Count of Genes") +
#     theme_minimal()
#   print(p)
# }

# # Determine overlap between SVM weights (considering a threshold)
# threshold <- 0.1
# for (i in 1:length(gene_counts)) {
#   for (j in (i+1):length(gene_counts)) {
#     signature_genes_model1 <- names(svm_weights_list[[as.character(gene_counts[i])]])[abs(svm_weights_list[[as.character(gene_counts[i])]]) > threshold]
#     signature_genes_model2 <- names(svm_weights_list[[as.character(gene_counts[j])]])[abs(svm_weights_list[[as.character(gene_counts[j])]]) > threshold]
#     overlap <- intersect(signature_genes_model1, signature_genes_model2)
#     
#     cat("Overlap between top", gene_counts[i], "and", gene_counts[j], "genes:", length(overlap), "\n")
#   }
# }


```

# BINARY CLASSIFICATION: This will run the classifier across only Control and Type2 for binary classification & AUC calculation.

```{r}
# Function to perform binary classification and return metrics
perform_binary_analysis <- function(scaled_matrix, labels) {
  # Prepare data for training
  data_train <- as.data.frame(scaled_matrix)
  data_train$label <- as.factor(labels)
  
  # Subset to only 'Control' and 'Type2' samples
  binary_data <- data_train[data_train$label %in% c('Control', 'Type2'), ]
  binary_data$label <- droplevels(binary_data$label)
  
  # Splitting data
  set.seed(123)
  data_split_binary <- initial_split(binary_data, prop = 0.8)
  train_data_binary <- training(data_split_binary)
  test_data_binary <- testing(data_split_binary)
  
  # Define and train SVM model
  svm_spec_binary <- svm_poly() %>%
    set_engine("kernlab") %>%
    set_mode("classification")
  
  svm_fit_binary <- svm_spec_binary %>%
    fit(label ~ ., data = train_data_binary)
  
  # Make predictions and evaluate
  predictions_binary <- predict(svm_fit_binary, test_data_binary)
  test_data_binary$predicted_label <- predictions_binary$.pred_class
  
  metrics_result_binary <- yardstick::metrics(test_data_binary, truth = label, estimate = predicted_label)
  print(paste("Results for top", ncol(scaled_matrix), "genes (Binary Case):"))
  print(metrics_result_binary)

  # Return metrics and SVM weights
  list(metrics = metrics_result_binary, svm_model = svm_fit_binary$fit)
  
}

# Main loop
gene_counts <- c(10, 100, 1000, 5000)#, 10000)
labels <- metadata$Group
accuracies_binary <- c()
svm_models_list_binary <- list()


for (n_genes in gene_counts) {
  scaled_matrix <- preprocess_data(expression_df, n_genes)
  analysis_result_binary <- perform_binary_analysis(scaled_matrix, labels)
  
  # Extract accuracy and store in vector
  acc <- analysis_result_binary$metrics$`.estimate`[analysis_result_binary$metrics$`.metric` == "accuracy"]
  accuracies_binary <- c(accuracies_binary, acc)
  
  # Store SVM weights
  svm_models_list_binary[[as.character(n_genes)]] <- analysis_result_binary$svm_model
  
  
}

df_binary <- data.frame(Number_of_Genes = gene_counts, Accuracy = accuracies_binary)
ggplot(df_binary, aes(x = Number_of_Genes, y = Accuracy)) +
  geom_line() +
  geom_point() +
  labs(title = "Binary Classification: Accuracy vs. Number of Genes", x = "Number of Genes", y = "Accuracy") +
  theme_minimal()
```
# BINARY CLASSIFICATION: looking at Gene Signatures
```{r}

# Accessing the SVM model from the list
svm_model_fit <- svm_models_list_binary[['5000']]

# Extract the support vectors
support_vectors <- svm_model_fit@xmatrix

# Number of support vectors
cat("Number of support vectors:", nrow(support_vectors[[1]]), "\n")

# Extract the coefficients (weights)
weights <- svm_model_fit@coef

# Extract all weights and combine them
all_weights <- unlist(weights)

# Visualizing the distribution of the weigghts
df_weights <- data.frame(Weight = all_weights)
ggplot(df_weights, aes(x=Weight)) +
  geom_histogram(bins=30, fill="blue", alpha=0.7) +
  labs(title="Distribution of SVM Weights") +
  theme_minimal()


```

```{r}

perform_auc_roc_analysis <- function(scaled_matrix, labels) {
  
  # Prepare data for training
  data_train <- as.data.frame(scaled_matrix) 
  data_train$label <- as.factor(labels)

  # Subset to only 'Control' and 'Type2' samples
  binary_data <- data_train[data_train$label %in% c('Control', 'Type2'), ]
  binary_data$label <- droplevels(binary_data$label)

  # Splitting data into training and testing sets (e.g., 80% training, 20% testing)
  set.seed(123) # Set seed for reproducibility
  data_split_binary <- initial_split(binary_data, prop = 0.8)
  train_data_binary <- training(data_split_binary)
  test_data_binary <- testing(data_split_binary)

  # Define SVM model specification for binary classification
  svm_spec_binary <- svm_poly() %>%
    set_engine("kernlab") %>%
    set_mode("classification")

  # Train the model on the binary data
  svm_fit_binary <- svm_spec_binary %>%
    fit(label ~ ., data = train_data_binary) 

  # Make probability predictions for binary test data
  probs_binary <- predict(svm_fit_binary, test_data_binary, type = "prob")

  # Bind the probabilities to the binary test data
  test_data_binary <- bind_cols(test_data_binary, probs_binary)

  # Compute AUC
  auc_result <- yardstick::roc_auc(test_data_binary, truth = label, .pred_Control)
  
  # Compute ROC curve data
  roc_data <- yardstick::roc_curve(test_data_binary, truth = label, .pred_Control)

  return(list(auc = auc_result, roc = roc_data))
}

gene_counts <- c(10, 100, 1000, 5000, 10000)
auc_results <- numeric()
roc_results <- list()  # This will store the roc data for each gene count

for (n_genes in gene_counts) {
  scaled_matrix <- preprocess_data(expression_df, n_genes)
  
  results <- perform_auc_roc_analysis(scaled_matrix, labels)
  
  # Store AUC and ROC curve data
  auc_results <- c(auc_results, results$auc$`.estimate`)
  roc_results[[as.character(n_genes)]] <- results$roc
}

# Plotting AUC vs. Number of Genes
df_auc <- data.frame(Number_of_Genes = gene_counts, AUC = auc_results)
ggplot(df_auc, aes(x = Number_of_Genes, y = AUC)) +
  geom_line() +
  geom_point() +
  labs(title = "AUC vs. Number of Genes", x = "Number of Genes", y = "AUC") +
  theme_minimal()

```

```{r}
library(ggplot2)
library(dplyr)

# Combine all ROC data frames into a single data frame
all_roc_data <- bind_rows(lapply(gene_counts, function(count) {
  df <- roc_results[[as.character(count)]]
  df$gene_count <- count
  return(df)
}), .id = "id")

# Now plot all ROC curves using a single ggplot call
plot <- ggplot(all_roc_data, aes(x = 1 - specificity, y = sensitivity, color = as.factor(gene_count))) +
  geom_line() +
  labs(title = "ROC Curves for Various Number of Genes", x = "False Positive Rate", y = "True Positive Rate") +
  theme_minimal() +
  coord_fixed(ratio = 1) + # To make sure the plot is square, as ROC plots usually are
  scale_color_discrete(name = "Number of Genes")

# Print the final plot with all ROC curves
print(plot)


```
